The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum

### 2014-03-20 cordillera-climate.tex
Typesetting: unit formatting

Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate. However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a~``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.

@@

As we aim to model glacial inception and growth of the Cordilleran ice sheet towards a~configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain. Each simulation starts from ice-free conditions and runs for 10\,ka, a~time interval representative of the rapid build-up of the last Cordilleran ice sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}. Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

@@

The Shallow Shelf Approximation (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}. SIA and SSA velocities are computed by finite difference methods on a~10\,km-resolution horizontal grid of 300 by 150 points (the modelling domain). Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a~depth of 1\,km. Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model. Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran ice sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.

@@

where $\phi$ is given in degrees and $z$ in meters above current sea level. This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-minute \citep{data:etopo1}. Sea level is lowered by 120\,m and basal topography responds to ice load following a~regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

@@

Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a~temperature-index (positive degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation when temperature is below 0\,\unit{\degree C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{\degree C}. Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{\degree C} in one year.

The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals. It accounts for temperature variability assuming a~normal distribution along a~central (input) value. The temperature standard deviation is a~constant model parameter and was assigned a~value of 3.07\,\unit{\degree C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}. The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature. This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse rate of 6\,\unit{\degree C\,km^{-1}} is used in all simulations. No corrections apply to precipitation changes with elevation.

@@

WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to the Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km-resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}. In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}. Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Figs.~\ref{fig:temp} and~\ref{fig:prec}).

@@

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10~ka using temperature offsets ranging from 0 to 10\,\unit{\degree C}. Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs. \ref{fig:ivolarea}-\ref{fig:cool05}).

Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}). For instance, using a~5\,\unit{\degree C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).

Under present climate (0\,\unit{\degree C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}). These results are fairly consistent with contemporary ice distribution, although both the 10~km horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a~meaningful detailed comparison. CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation. At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extents are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}). The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.

To contrast the effect of the six different input climatologies, we plot simulations using a~5\,\unit{\degree C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a~large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a~more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations. These new simulations use a~“hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A~single temperature offset of 5\,\unit{\degree C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{\degree C} run (Fig.~\ref{fig:cool05}, top-left panel). For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}). However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).

@@

We compare modelled ice sheet geometries to a~reconstruction of the ice margin at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}). This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$. This corresponds to temperature offset values of 8\,\unit{\degree C} (WorldClim), 6\,\unit{\degree C} (ERA-Interim) and 7\,\unit{\degree C} (NARR). These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

Secondly, our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean (Fig.~\ref{fig:best}). These regions experience a~marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}). However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land. This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.

@@

A~key assumption for the present study is our choice of 10\,ka for the duration of all our simulations. Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,ka, they provide no lower bound to this value.

To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{\degree C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$.

@@

% tab:reanalyses
\begin{table*}[t]
	\caption{Characteristics of reanalysis climatologies used to inform the mass balance component of PISM.}
	\label{tab:reanalyses}
	\vskip4mm
	\centering
	\begin{tabular}{lllll}
		\tophline
		Reanalysis& Spatial coverage& Averaging period& Resolution& Description\\
		\middlehline
		NCEP/NCAR&  global&     1981 -- 2010& 1.875\degree& \citet{data:ncar}\\
		ERA-Interim&global&     1979 -- 2011& 1.000\degree& \citet{data:erai}\\
		CFSR&       global&     1979 -- 2010& 0.325\degree& \citet{data:cfsr}\\
		NARR&       North America& 1979 -- 2000& 32\,km& \citet{data:narr}\\
		\bottomhline
	\end{tabular}
\end{table*}

% fig:locmap
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=8cm]{cordillera-climate-locmap}
	\end{center}
	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,$^{14}$C\,ka\,BP (16.8\,cal\,ka\,BP) \citep{dyke-2004}. While this age denotes the LIS after retreat from its LGM, it closely corresponds to the LGM extent of most of the Cordilleran ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010}. The rectangular box denotes the modelling domain of the Cordilleran ice sheet of 1500 by 3000\,km. Major mountain ranges covered by the ice sheet include the Wrangell and St.-Elias mountains (W-SE), the Selwyn and MacKenzie mountains (S-MK), the Coast Mountains and the Rocky Mountains. The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth} and was assembled with GRASS~GIS \citep{soft:grass}.}
	\label{fig:locmap}
\end{figure}

@@

% fig:ivolarea
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-ivolarea}
	\end{center}
	\caption{Total glaciated area and ice volume after 10\,ka as a~function of temperature offset for each climate forcing used.}
	\label{fig:ivolarea}
\end{figure}

% fig:extent
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-extent}
	\end{center}
	\caption{Extent of ice cover after 10\,ka as a~function of applied temperature offsets for each climate forcing used. A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (black line).}
	\label{fig:extent}
\end{figure*}

% fig:cool05
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-cool05}
	\end{center}
	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka under a~climate 5\,\unit{\degree C} colder than present for each climate forcing used.}
	\label{fig:cool05}
\end{figure*}

@@

% fig:biatm
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatm}
	\end{center}
	\caption{Ice surface topography (1 km contours) and velocity (\unit{m\,a^{-1}}) after 10\,ka using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row). In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data. Each simulation uses a~5\,\unit{\degree C} offset for comparison with figure~\ref{fig:cool05}}
	\label{fig:biatm}
\end{figure*}

% fig:biatmbars
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatmbars}
	\end{center}
	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{\degree C} offset simulation. Corresponding ice sheet geometries are presented in figures~\ref{fig:cool05} and~\ref{fig:biatm}}
	\label{fig:biatmbars}
\end{figure}

% fig:best
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-best}
	\end{center}
	\caption{Ice surface topography (1\,km contours) after 10\,ka using temperature offsets resulting in glaciated areas of circa $2\,\times10^6\,\unit{km^2}$. A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}
	\label{fig:best}
\end{figure*}

% fig:durationstack
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-durationstack}
	\end{center}
	\caption{Left panel: model sensitivity to simulation length. Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{\degree C}, solid lines corresponding to values from 7 to 11\,\unit{\degree C}. Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{\degree C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2\,\times10^6\,\unit{km^2}$. A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading). Note that shorter (and colder) simulations lead to more restrictive glaciation of the continental eastern margin but further ice extent in the maritime south-western part of the modelling domain.}
	\label{fig:durationstack}
\end{figure}

>>>>

Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project \citep[PMIP;][]{joussaume-taylor-1995} provide perhaps a~more reasonable representation of past climate. However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a~``glacial index'' weighting function derived from ice core \chem{\delta^{18}O} records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.

@@

As we aim to model glacial inception and growth of the Cordilleran ice sheet towards a~configuration as last attained during the LGM, we mimic palaeo-climatic conditions by applying constant temperature offsets homogeneously over the modelling domain. Each simulation starts from ice-free conditions and runs for 10\,kyr, a~time interval representative of the rapid build-up of the last Cordilleran ice sheet from nearly ice-free to full glacial conditions \citep{clague-1989,stroeven-etal-2010}. Our simulations are performed in parallel on 32 cores at the Swedish National Supercomputing Center.

@@

The Shallow Shelf Approximation (SSA) is used as a~``sliding law'' for the Shallow Ice Approximation (SIA) \citep{bueler-brown-2009,winkelmann-etal-2011}. SIA and SSA velocities are computed by finite difference methods on a~10\,km resolution horizontal grid of 300 by 150 points (the modelling domain). Ice softness depends on temperature and water content through an enthalpy formulation \citep{aschwanden-blatter-2009,aschwanden-etal-2012}. Enthalpy is computed in three dimensions in up to 51 irregularly spaced layers within the ice, and temperature is additionally computed in 11 regularly spaced layers in bedrock to a~depth of 1\,km. Surface air temperature from the climate forcing provides the upper boundary condition to the ice enthalpy model, and a~geothermal heat flux of 70\,\unit{mW\,m^{-2}} provides the lower boundary condition to the bedrock thermal model. Although this uniform value does not account for the high spatial geothermal variability in the region, it is on average representative of available heat flow measurements \citep{artemieva-mooney-2001,blackwell-richards-2004}. Whereas spatial variations in the geothermal heat flux may significantly affect basal conditions \citep{pattyn-2010}, this study focuses on reconstructing the former extent of the Cordilleran ice sheet, on which it has relatively little effect \citep{rogozhina-etal-2012}.

@@

where $\phi$ is given in degrees and $z$ in meters above current sea level. This model thus accounts for a~weakening of till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a~resolution of 1\,arc-min \citep{data:etopo1}. Sea level is lowered by 120\,m and basal topography responds to ice load following a~regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

@@

Ice surface accumulation and ablation are computed from monthly mean surface air temperature and monthly precipitation by a~temperature-index (positive degree-day) model \citep{hock-2003}. Ice accumulation is equal to precipitation when temperature is below 0\,\unit{{\degree}C}, and decreases to zero linearly with temperature between 0 and 2\,\unit{{\degree}C}. Ice ablation is computed from the number of positive degree-days, defined as the integral of temperatures above 0\,\unit{{\degree}C} in one year.

The positive degree-day integral \citep{calov-greve-2005} is numerically approximated using week-long sub-intervals. It accounts for temperature variability assuming a~normal distribution along a~central (input) value. The temperature standard deviation is a~constant model parameter and was assigned a~value of 3.07\,\unit{{\degree}C}, which corresponds to the mean summer (JJA), model domain-averaged monthly standard deviation of daily mean temperature from monthly mean temperature, as computed from North American Regional Reanalysis data \citep{data:narr} in a~manner similar to \citet{seguinot-2013}. The ablation model incorporates degree-day factors of 3.04\,\unit{mm\,K^{-1}\,day^{-1}} for snow and 4.59\,\unit{mm\,K^{-1}\,day^{-1}} for ice, as derived from mass-balance measurements on contemporary glaciers from the Coast Mountains and Rocky Mountains in British Columbia \citep{shea-etal-2009}.

@@

Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature. This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all simulations. No corrections apply to precipitation changes with elevation.

@@

WorldClim, NCEP/NCAR, ERA-Interim, CFSR and NARR climatologies were re-projected to the Canadian Atlas Lambert conformal conic projection (EPSG code~3978) and bilinearly interpolated to the 10\,km resolution model grid using the module r.proj from GRASS~GIS \citep{soft:grass}. In addition, WorldClim data was extrapolated to cover grid points in the ocean using the nearest-neighbour algorithm \citep{soft:scipy}. Note that here we choose to present original rather than interpolated data, in order to highlight differences between datasets related to spatial resolution (Figs.~\ref{fig:temp} and~\ref{fig:prec}).

@@

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from 0 to 10\,\unit{{\degree}C}. Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs. \ref{fig:ivolarea}-\ref{fig:cool05}).

Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}). For instance, using a~5\,\unit{{\degree}C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).

Under present climate (0\,\unit{{\degree}C} offset), WorldClim, ERA-Interim and NARR forcing produce local ice caps in areas presently glaciated (Fig.~\ref{fig:extent}). These results are fairly consistent with contemporary ice distribution, although both the 10\,\unit{km} horizontal resolution and the occurrence of glaciers and ice caps surface topographies in the ETOPO1 dataset used for basal topography prevent a~meaningful detailed comparison. CFSR forcing, on the other hand, yields continuous ice cover over the northern Coast Mountains, and NCEP/NCAR forcing results in ice sheet growth, both of which are largely inconsistent with the present state of glaciation. At the opposite side of the temperature offset range, it can be noted that both the NCEP/NCAR and CFSR forcing produce oversized ice sheets whose extents are primarily restricted by the domain boundaries (Fig.~\ref{fig:extent}). The smoothing of CFSR precipitation yields an insignificant effect on the resulting ice sheet geometry.

To contrast the effect of the six different input climatologies, we plot simulations using a~5\,\unit{{\degree}C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a~large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a~more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations. These new simulations use a~“hybrid” climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A~single temperature offset of 5\,\unit{{\degree}C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel). For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}). However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap}--\ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap}--\ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).

@@

We compare modelled ice sheet geometries to a~reconstruction of the ice margin at 14\,\unit{\chem{^{14}C}\,kyr\,BP} (16.8\,cal\,kyr\,BP) by \citet{dyke-2004}, based on glacial geomorphology and radiocarbon dating (Fig.~\ref{fig:locmap}). This corresponds to the LGM extent of most of the Cordilleran ice sheet, which occurred later than for the Laurentide ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010,stroeven-etal-inpress}.

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$. This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR). These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

Secondly, our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean (Fig.~\ref{fig:best}). These regions experience a~marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}). However in our simulations, sea-level is lowered by 120\,\unit{m}, turning large parts of the low-sloping continental shelf into land. This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.

@@

A~key assumption for the present study is our choice of 10\,kyr for the duration of all our simulations. Although geomorphological data shows that the period of Cordilleran ice sheet inception, from nearly ice-free to full glacial conditions, could not have been much longer than 10\,kyr, they provide no lower bound to this value.

To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{{\degree}C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.

@@

% tab:reanalyses
\begin{table*}[t]
	\caption{Characteristics of reanalysis climatologies used to inform the mass balance component of PISM.}
	\label{tab:reanalyses}
	\vskip4mm
	\centering
	\begin{tabular}{lllll}
		\tophline
		Reanalysis& Spatial coverage& Averaging period& Resolution& Description\\
		\middlehline
    NCEP/NCAR   & global        & 1981--2010 & 1.875{\degree} & \citet{data:ncar} \\
    ERA-Interim & global        & 1979--2011 & 1.000{\degree} & \citet{data:erai} \\
    CFSR        & global        & 1979--2010 & 0.325{\degree} & \citet{data:cfsr} \\
    NARR        & North America & 1979--2000 & 32\,km       & \citet{data:narr} \\
		\bottomhline
	\end{tabular}
\end{table*}

% fig:locmap
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics[width=8cm]{cordillera-climate-locmap}
	\end{center}
	\caption{Shaded relief map of northern North America with the extent of the Cordilleran (CIS), Laurentide (LIS) and Innuitian (IIS) ice sheets at 14\,\unit{\chem{^{14}C}\,kyr\,BP} (16.8\,cal\,kyr\,BP) \citep{dyke-2004}. While this age denotes the LIS after retreat from its LGM, it closely corresponds to the LGM extent of most of the Cordilleran ice sheet \citep{porter-swanson-1998,dyke-2004,stroeven-etal-2010}. The rectangular box denotes the modelling domain of the Cordilleran ice sheet of 1500 by 3000\,km. Major mountain ranges covered by the ice sheet include the Wrangell and St.-Elias mountains (W-SE), the Selwyn and MacKenzie mountains (S-MK), the Coast Mountains and the Rocky Mountains. The background map consists of ETOPO1 \citep{data:etopo1} and Natural Earth Data \citep{data:naturalearth} and was assembled with GRASS~GIS \citep{soft:grass}.}
	\label{fig:locmap}
\end{figure}

@@

% fig:ivolarea
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-ivolarea}
	\end{center}
	\caption{Total glaciated area and ice volume after 10\,kyr as a~function of temperature offset for each climate forcing used.}
	\label{fig:ivolarea}
\end{figure}

% fig:extent
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-extent}
	\end{center}
	\caption{Extent of ice cover after 10\,kyr as a~function of applied temperature offsets for each climate forcing used. A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (black line).}
	\label{fig:extent}
\end{figure*}

% fig:cool05
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-cool05}
	\end{center}
	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a~climate 5\,\unit{{\degree}C} colder than present for each climate forcing used.}
	\label{fig:cool05}
\end{figure*}

@@

% fig:biatm
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatm}
	\end{center}
	\caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using “hybrid” climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row). In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data. Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with figure~\ref{fig:cool05}}
	\label{fig:biatm}
\end{figure*}

% fig:biatmbars
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-biatmbars}
	\end{center}
	\caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{{\degree}C} offset simulation. Corresponding ice sheet geometries are presented in figures~\ref{fig:cool05} and~\ref{fig:biatm}}
	\label{fig:biatmbars}
\end{figure}

% fig:best
\begin{figure*}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-best}
	\end{center}
	\caption{Ice surface topography (1\,km contours) after 10\,kyr using temperature offsets resulting in glaciated areas of circa $2 \times 10^6\,\unit{km^2}$. A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (blue line).}
	\label{fig:best}
\end{figure*}

% fig:durationstack
\begin{figure}[t]
	\vspace*{2mm}
	\begin{center}
		\includegraphics{cordillera-climate-durationstack}
	\end{center}
	\caption{Left panel: model sensitivity to simulation length. Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{{\degree}C}, solid lines corresponding to values from 7 to 11\,\unit{{\degree}C}. Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{{\degree}C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2 \times 10^6\,\unit{km^2}$. A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading). Note that shorter (and colder) simulations lead to more restrictive glaciation of the continental eastern margin but further ice extent in the maritime south-western part of the modelling domain.}
	\label{fig:durationstack}
\end{figure}
